### 2020-11-13 docs/manuscript/PLOS-MSc-Thesis.tex
correct typos \noindent and equations

\subsection*{Background}
A central problem in developmental and synthetic biology is to understand the 
mechanisms by which cells in a tissue or in a petri dish process external cues 
and transform such information into a coherent response, e.g. a terminal 
differentiation state. It was long believed that this type of positional 
information could be entirely attributed to a gradient of concentration of a 
certain signaling molecule (i.e. a morphogen). However, advances in experimental 
methodologies and computer modeling have demonstrated the crucial role of the 
dynamics of a cell's gene regulatory network (GRN) in decoding the information 
carried by the morphogen, which is eventually translated into a spatial 
pattern. This morphogen interpretation mechanism has gained a lot of attention 
in systems biology as a tractable system to investigate the emergent properties 
of complex genotype-phenotype maps.

@@

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Systems biology is growing very quickly thanks to the advances in computing and
mathematical modeling of biological systems, and has proven to be a powerful tool that
can help us to understand complex biological processes such as pattern formation and
development. In this article we use an \textit{in silico} approach in order to find and study
the design space of three node Gene Regulatory Networks that are able to produce a 
striped pattern of gene expression.

@@

 \begin{eqnarray}
  \frac{d[G]^i_n}{dt}
  = g(u^i) + D_i \cdot [ ([G]^i_{n-1}-[G]^i_n) +([G]^i_{n+1}-[G]^i_n)]-\delta_i 
  \cdot [ G ]^i_n ,
  \label{ec:modelo-gral}
 \end{eqnarray}

in which $[G]^{i}_{n}$ is the concentration of the i-th gene in the \emph{n}-th 
cell ($[G]^{i}_{n} ≥ 0$), $g(u^i)$ is a function describing the relationship 
between the interactions on the \emph{i}-th gene and its expression 
Eq~(\ref{ec:input-func}) and described in more detail below. $D_i$ and 
$\delta_i$ are the diffusion and the degradation rate of \emph{i}-th gene 
product, respectively.\\

The input function is a sigmoid function described by Eq~\ref{ec:input-func}.

\begin{equation}
 g(u^i) = \frac{1}{1 + \exp(a - b \cdot u^i)},
 \label{ec:input-func}
\end{equation}

in which $a$ is the sigmoid steepness, equal to 5; $a/b$ is the threshold value,
set to 1 for all simulations, and $u^i$ is the following equation:

\begin{equation}
 u^i = \sum_j w_{ij} \cdot [G]^j_n + w_{im} \cdot [M]_n.
 \label{ec:mat-sum}
\end{equation}

@@

The morphogen concentration along the morphogenetic field is described by 
Eq~\ref{ec:morph}.

\begin{equation}
 M = A_0 \cdot \exp(-c/h),
 \label{ec:morph}
\end{equation}

were $A_0$ is the concentration of the morphogen in the position zero of the 
morphogenetic field and was set to 1 in our experiments; $c$ is the cell index, 
defined as the ratio between the \emph{n}-th cell from left to right and the 
total number of cells, and $h$ is a decay parameter, whose value in our model 
was set to 0.4.

@@

 \item{\bf{Numerically solve the dynamical system for the GRN from t=0 until 
t=500.} We integrated the differential equation using the function NDSolve from 
Wolfram Mathematica 11 with the option “EquationSimplification” and the 
“Residual” simplification method.}

 \item{\bf Evaluate fitness by comparing the GRN phenotype with the optimal 
 phenotype (Fig~\ref{fig:exp-pattern}).} In order to compute the fitness 
 function for each GRN we used three different filters. The first filter 
 assesses whether the expression profile of the output node reaches a 
 quasi-steady state:

 \begin{equation}
  S_{filter} = \left( \frac{1}{I \cdot N}\right) \sqrt{\sum_{i=1}^{I=3}
  \sum_{n=1}^{N=30} ([G]_n^i(t=500) - [G]^i_n(t=250))^2}.
 \end{equation}

@@

 The second filter measures the spatial heterogeneity in the field:

 \begin{equation}
  P_{filter} = \left( \frac{1}{I \cdot N} \right) \sqrt{ \sum_{i=3}^{I=3}
  \sum_{n=1}^{N=30} \left( [G]^i_n - \langle [G]^i_n \rangle \right)^2 },
 \end{equation}

@@

 The third filter relates the Manhattan Distance ($D_{obs}$) between an
 expression pattern and the optimal pattern with the maximum distance achievable
 ($D_{max}$):

 \begin{equation}
  \mathit{PF}_{eff} = 1 - \frac{D_{obs}}{D_{max}}.
 \end{equation}

@@

 The filters mentioned above are integrated in the following functions that 
 asses the quality of a given phenotype:

 \begin{equation}
  Q_1(P_{filter}) = \left( \frac{ {P_{filter}}^{10} }{ {P_{filter}}^{10} + 
  0.1^{10} } \right),
 \end{equation}

 \begin{equation}
  Q_2(S_{filter}) = \left( \frac{ 0.1^2 }{ {S_{filter}}^2 + 0.1^2 } \right).
 \end{equation}

 Finally, these quality functions were used to compute the fitness score with
 the following fitness function:

 \begin{equation}
  F = \mathit{PF_{eff}} \cdot Q_1(P_{filter}) \cdot Q_2(S_{filter}).
 \end{equation}

@@

The neutral network was obtained by generating all possible isomorphs of a 
network topology and then calculating the minimal Hamming distance between these
isomorphs and other network topologies. The Hamming distance between two 
topologies was calculated using the following equation:

\begin{equation}
 D_H(W^A, W^B) = \sum_{ij} \mid sgn(w_{ij}^A) - sgn(w^B_{ij}) \mid .
\end{equation}

@@

In order to evaluate the robustness of the different topologies to perturbations
in parameters of interactions between genes we took a GRN belonging to each one 
of the network topologies and we modified each interaction independently by 
increasing and decreasing its value in a 20\%. For topologies with 3 or more 
GRNs we chose the most similar to the mean configuration for that topology.\\

@@

We calculated the Shannon entropy Eq~(\ref{ec:shannon}) of the network 
topologies distribution in order to test if this distribution could be obtained 
by chance with a probability greater than 0.05. In order to obtain a 95\% 
confidence interval, we generated 30 samples of a set of 2061 random matrices 
and calculated the Shannon entropy in bits for each sample and tested these data
for normality.

\begin{equation}
 H = -\sum p(x) \cdot \log_{2}p(x).
 \label{ec:shannon}
\end{equation}

@@

As a measure of network complexity, we used the complexity index based on 
Shannon entropy proposed by Bonchev \& Rovray \cite{D.2005}. We calculated this 
index using the following equation:

\begin{equation}
 I_{vd} = \sum_{i=1}^V a_i \cdot \log_{2} a_i,
\end{equation}

being $a_i$ the degree of \emph{i}-th node and $V$ the number of nodes in the 
network. The network descriptors were calculated using built-in functions from 
Wolfram Mathematica software.\\

@@

We found that the I3-FFL is one of the network motifs that appears more
frequently in these GRNs, as has also been reported by Cotterell \& Sharpe
\cite{Cotterell2010}, and has been found in transcriptional networks involved 
in cell fate definition \cite{Li2019}. Moreover, our results indicate that this 
is a robust motif that can be usefull for construction of synthetic gene 
regulatory networks that are exposed to stochastic fluctuations in the 
environment and mutations.

>>>>

\subsection*{Background}
A central problem in developmental and synthetic biology is to understand the 
mechanisms by which cells in a tissue or in a Petri dish process external cues 
and transform such information into a coherent response, e.g. a terminal 
differentiation state. It was long believed that this type of positional 
information could be entirely attributed to a gradient of concentration of a 
certain signaling molecule (i.e. a morphogen). However, advances in experimental 
methodologies and computer modeling have demonstrated the crucial role of the 
dynamics of a cell's gene regulatory network (GRN) in decoding the information 
carried by the morphogen, which is eventually translated into a spatial 
pattern. This morphogen interpretation mechanism has gained a lot of attention 
in systems biology as a tractable system to investigate the emergent properties 
of complex genotype-phenotype maps.

@@

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Systems biology is a very fast growing field thanks to the advances in computing and
mathematical modeling of biological systems, and has proven to be a powerful tool that
can help us to understand complex biological processes such as pattern formation in
development. In this article we use an \textit{in silico} approach in order to find and study
the design space of three node Gene Regulatory Networks that are able to produce a 
striped pattern of gene expression.

@@

 \begin{eqnarray}
  \frac{d[G]^i_n}{dt}
  = g(u^i) + D_i \cdot [ ([G]^i_{n-1}-[G]^i_n) +([G]^i_{n+1}-[G]^i_n)]-\delta_i 
  \cdot [ G ]^i_n ,
  \label{ec:modelo-gral}
 \end{eqnarray}
\noindent
in which $[G]^{i}_{n}$ is the concentration of the i-th gene in the \emph{n}-th 
cell ($[G]^{i}_{n} ≥ 0$), $g(u^i)$ is a function describing the relationship 
between the interactions on the \emph{i}-th gene and its expression 
Eq~(\ref{ec:input-func}) and described in more detail below. $D_i$ and 
$\delta_i$ are the diffusion and the degradation rate of \emph{i}-th gene 
product, respectively.\\

The input function is a sigmoid function described by Eq~\ref{ec:input-func}.
\begin{equation}
 g(u^i) = \frac{1}{1 + \exp(a - b \cdot u^i)},
 \label{ec:input-func}
\end{equation}
\noindent
in which $a$ is the sigmoid steepness, equal to 5; $a/b$ is the threshold value,
set to 1 for all simulations, and $u^i$ is the following equation:
\begin{equation}
 u^i = \sum_j w_{ij} \cdot [G]^j_n + w_{im} \cdot [M]_n.
 \label{ec:mat-sum}
\end{equation}

@@

The morphogen concentration along the morphogenetic field is described by 
Eq~\ref{ec:morph}.
\begin{equation}
 M = A_0 \cdot \exp(-c/h),
 \label{ec:morph}
\end{equation}
\noindent
were $A_0$ is the concentration of the morphogen in the position zero of the 
morphogenetic field and was set to 1 in our experiments; $c$ is the cell index, 
defined as the ratio between the \emph{n}-th cell from left to right and the 
total number of cells, and $h$ is a decay parameter, whose value in our model 
was set to 0.4.

@@

 \item{\bf Numerically solve the dynamical system for the GRN from t=0 until 
 t=500.} We integrated the differential equation using the function NDSolve from 
 Wolfram Mathematica 11 with the option “EquationSimplification” and the 
 “Residual” simplification method.

 \item{\bf Evaluate fitness by comparing the GRN phenotype with the optimal 
 phenotype (Fig~\ref{fig:exp-pattern}).} In order to compute the fitness 
 function for each GRN we used three different filters. The first filter 
 assesses whether the expression profile of the output node reaches a 
 quasi-steady state:
 \begin{equation}
  S_{filter} = \left( \frac{1}{I \cdot N}\right) \sqrt{\sum_{i=1}^{I=3}
  \sum_{n=1}^{N=30} ([G]_n^i(t=500) - [G]^i_n(t=250))^2}.
 \end{equation}

@@

 The second filter measures the spatial heterogeneity in the field:
 \begin{equation}
  P_{filter} = \left( \frac{1}{I \cdot N} \right) \sqrt{ \sum_{i=3}^{I=3}
  \sum_{n=1}^{N=30} \left( [G]^i_n - \langle [G]^i_n \rangle \right)^2 },
 \end{equation}

@@

 The third filter relates the Manhattan Distance ($D_{obs}$) between an
 expression pattern and the optimal pattern with the maximum distance achievable
 ($D_{max}$):
 \begin{equation}
  \mathit{PF}_{eff} = 1 - \frac{D_{obs}}{D_{max}}.
 \end{equation}

@@

 The filters mentioned above are integrated in the following functions that 
 evaluates the  quality of a given phenotype:

 \begin{equation}
  Q_1(P_{filter}) = \left( \frac{ {P_{filter}}^{10} }{ {P_{filter}}^{10} + 
  0.1^{10} } \right),
 \end{equation}
 \begin{equation}
  Q_2(S_{filter}) = \left( \frac{ 0.1^2 }{ {S_{filter}}^2 + 0.1^2 } \right).
 \end{equation}

 Finally, these quality functions were used to compute the fitness score with
 the following fitness function:
 \begin{equation}
  F = \mathit{PF_{eff}} \cdot Q_1(P_{filter}) \cdot Q_2(S_{filter}).
 \end{equation}

@@

The neutral network was obtained by generating all possible isomorphs of a 
network topology and then calculating the minimal Hamming distance between these
isomorphs and other network topologies. The Hamming distance between two 
topologies was calculated using the following equation:
\begin{equation}
 D_H(W^A, W^B) = \sum_{ij} \mid sgn(w_{ij}^A) - sgn(w^B_{ij}) \mid .
\end{equation}

@@

In order to evaluate the robustness of the different topologies to perturbations
in parameters of interactions between genes we took a GRN belonging to each one 
of the network topologies and we modified each interaction independently by 
increasing and decreasing its value by 20\%. For topologies with 3 or more 
GRNs we chose the most similar to the mean configuration for that topology.\\

@@

We calculated the Shannon entropy Eq~(\ref{ec:shannon}) of the network 
topologies distribution in order to test if this distribution could be obtained 
by chance with a probability greater than 0.05. In order to obtain a 95\% 
confidence interval, we generated 30 samples of a set of 2061 random matrices 
and calculated the Shannon entropy in bits for each sample and tested these data
for normality.
\begin{equation}
 H = -\sum p(x) \cdot \log_{2}p(x).
 \label{ec:shannon}
\end{equation}

@@

As a measure of network complexity, we used the complexity index based on 
Shannon entropy proposed by Bonchev \& Rovray \cite{D.2005}. We calculated this 
index using the following equation:
\begin{equation}
 I_{vd} = \sum_{i=1}^V a_i \cdot \log_{2} a_i,
\end{equation}
\noindent
being $a_i$ the degree of \emph{i}-th node and $V$ the number of nodes in the 
network. The network descriptors were calculated using built-in functions from 
Wolfram Mathematica software.\\

@@

We found that the I3-FFL is one of the network motifs that appears more
frequently in these GRNs, as has also been reported by Cotterell \& Sharpe
\cite{Cotterell2010}, and has been found in transcriptional networks involved 
in cell fate definition \cite{Li2019}. Moreover, our results indicate that this 
is a robust motif that can be useful for construction of synthetic gene 
regulatory networks that are exposed to stochastic fluctuations in the 
environment and mutations.
